---
title: "Marketing Performance Analysis"
author: "Eetu Tammi"
format: html
editor: visual
---

## Introduction

This project analyzes the relationship between marketing spending and weekly revenue using a simulated marketing dataset.

The analysis focuses on five marketing channels: TV, OOH, Print, Facebook, and Search. It also considers competitor sales and seasonality.

The first research question is:

**How is marketing spending associated with weekly revenue?**

The analysis examines marketing spend by channel, correlations between marketing activity and revenue, seasonality, and a multiple linear regression model.

The second research question is:

**How well can revenue be predicted using marketing activity, competitor sales, and seasonality?**

The model is evaluated using a chronological train/test split, with observations before 2019 used for training and 2019 observations used as the test set.

The third research question is:

**Do marketing activities from previous weeks improve revenue predictions?**

One- and two-week lags are added to each marketing channel and compared with the simpler baseline model using MAE, RMSE, and MAPE.

## Data

The analysis uses Meta's simulated weekly marketing dataset.

The dataset contains weekly observations of:

- revenue
- TV spend
- OOH spend
- Print spend
- Facebook spend
- Search spend
- competitor sales
- events
- newsletter activity

The analysis focuses on the five paid marketing channels, competitor sales, and calendar month.

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(scales)
library(broom)

load("data/dt_simulated_weekly.RData")

df <- dt_simulated_weekly %>%
  arrange(DATE) %>%
  mutate(
    month = factor(month(DATE))
  )

marketing_channels <- c(
  "tv_S",
  "ooh_S",
  "print_S",
  "facebook_S",
  "search_S"
)

marketing_labels <- c(
  tv_S = "TV",
  ooh_S = "OOH",
  print_S = "Print",
  facebook_S = "Facebook",
  search_S = "Search"
)
```

## Data Preparation

The data contains 208 weekly observations covering the period from November 2015 to November 2019.

The analysis uses the following marketing variables:

- `tv_S`
- `ooh_S`
- `print_S`
- `facebook_S`
- `search_S`

Competitor sales are included as a control variable because they represent an external business factor that is strongly associated with revenue.

Calendar month is included to account for recurring seasonal patterns.

```{r, echo=FALSE, message=FALSE}
dataset_summary <- tibble(
  observations = nrow(df),
  variables = ncol(df),
  start_date = min(df$DATE),
  end_date = max(df$DATE),
  missing_values = sum(is.na(df)),
  duplicate_rows = sum(duplicated(df))
)

dataset_summary
```

The dataset contains **208 weekly observations** and no missing values or duplicate rows.

## Exploratory Analysis

### Revenue Over Time

The first step is to examine how revenue changes over time.

```{r, echo=FALSE}
ggplot(
  df,
  aes(
    x = DATE,
    y = revenue
  )
) +
  geom_line() +
  scale_y_continuous(
    labels = comma
  ) +
  labs(
    x = "Date",
    y = "Revenue"
  ) +
  theme_minimal()
```

Revenue shows substantial variation over time and a recurring seasonal pattern.

### Marketing Spend by Channel

The total marketing spend is calculated for each channel.

```{r, echo=FALSE}
spend_summary <- df %>%
  summarise(
    across(
      all_of(marketing_channels),
      sum
    )
  ) %>%
  pivot_longer(
    everything(),
    names_to = "channel",
    values_to = "total_spend"
  ) %>%
  mutate(
    channel = recode(
      channel,
      !!!marketing_labels
    ),
    share = total_spend / sum(total_spend) * 100
  ) %>%
  arrange(
    desc(total_spend)
  )

spend_summary
```

OOH accounts for the largest share of marketing spending, followed by TV and Search.

However, the amount spent on a channel does not by itself indicate how effective that channel is.

```{r, echo=FALSE}
ggplot(
  spend_summary,
  aes(
    x = reorder(channel, total_spend),
    y = total_spend
  )
) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(
    labels = comma
  ) +
  labs(
    x = NULL,
    y = "Total marketing spend"
  ) +
  theme_minimal()
```

### Marketing Spend and Revenue

The relationship between total marketing spending and revenue is examined next.

```{r, echo=FALSE}
df <- df %>%
  mutate(
    total_marketing_spend =
      tv_S +
      ooh_S +
      print_S +
      facebook_S +
      search_S
  )

ggplot(
  df,
  aes(
    x = total_marketing_spend,
    y = revenue
  )
) +
  geom_point() +
  geom_smooth(
    method = "lm",
    se = FALSE
  ) +
  scale_x_continuous(
    labels = comma
  ) +
  scale_y_continuous(
    labels = comma
  ) +
  labs(
    x = "Total marketing spend",
    y = "Revenue"
  ) +
  theme_minimal()
```

The relationship is positive, but this simple relationship does not account for seasonality, competitor activity, or the simultaneous use of different marketing channels.

### Correlation Between Marketing Channels and Revenue

```{r, echo=FALSE}
correlation_variables <- c(
  "revenue",
  marketing_channels,
  "competitor_sales_B"
)

correlation_matrix <- cor(
  df %>%
    select(
      all_of(correlation_variables)
    )
)

round(
  correlation_matrix,
  3
)
```

The correlations provide an initial indication of which variables move together with revenue.

The strongest individual marketing correlations with revenue are observed for Search and TV. Competitor sales have a particularly strong correlation with revenue.

### Competitor Sales

```{r, echo=FALSE}
competitor_correlation <- cor(
  df$revenue,
  df$competitor_sales_B
)

competitor_correlation
```

The correlation between competitor sales and revenue is approximately **0.916**.

```{r, echo=FALSE}
ggplot(
  df,
  aes(
    x = competitor_sales_B,
    y = revenue
  )
) +
  geom_point() +
  geom_smooth(
    method = "lm",
    se = FALSE
  ) +
  scale_x_continuous(
    labels = comma
  ) +
  scale_y_continuous(
    labels = comma
  ) +
  labs(
    x = "Competitor sales",
    y = "Revenue"
  ) +
  theme_minimal()
```

The strong relationship suggests that competitor activity is an important control variable when estimating the relationship between marketing spending and revenue.

### Seasonality

Average revenue is compared across calendar months.

```{r, echo=FALSE}
monthly_revenue <- df %>%
  group_by(
    month
  ) %>%
  summarise(
    average_revenue = mean(revenue),
    .groups = "drop"
  )

monthly_revenue
```

```{r, echo=FALSE}
ggplot(
  monthly_revenue,
  aes(
    x = month,
    y = average_revenue
  )
) +
  geom_col() +
  scale_y_continuous(
    labels = comma
  ) +
  labs(
    x = "Month",
    y = "Average revenue"
  ) +
  theme_minimal()
```

Revenue varies substantially across months. The average revenue is considerably higher in the autumn and early winter than during the summer.

This seasonal variation is therefore included in the regression model.

## Research Question 1: Marketing Spending and Revenue

### Question

**How is marketing spending associated with weekly revenue after controlling for competitor activity and seasonality?**

A multiple linear regression model is used to estimate the relationship between revenue and the five marketing channels.

The model also includes competitor sales and calendar month as control variables.

```{r, echo=FALSE}
model_baseline <- lm(
  revenue ~
    tv_S +
    ooh_S +
    print_S +
    facebook_S +
    search_S +
    competitor_sales_B +
    month,
  data = df
)

summary(model_baseline)
```

### Regression Results

```{r, echo=FALSE}
baseline_coefficients <- tidy(
  model_baseline
) %>%
  filter(
    term != "(Intercept)"
  )

baseline_coefficients
```

The model has an adjusted R-squared of approximately **0.87**, indicating that the included variables explain a large share of the variation in revenue in the sample.

TV has a positive and statistically significant coefficient. Print is also statistically significant in the training model.

OOH, Facebook, and Search are not statistically significant after controlling for the other variables and seasonality.

These coefficients should be interpreted as **associations, not causal effects**.

A statistically insignificant coefficient does not prove that a marketing channel has no effect on revenue.

### Marketing Coefficients

```{r, echo=FALSE}
marketing_coefficients <- tidy(
  model_baseline
) %>%
  filter(
    term %in% marketing_channels
  ) %>%
  mutate(
    channel = recode(
      term,
      !!!marketing_labels
    )
  )

ggplot(
  marketing_coefficients,
  aes(
    x = reorder(channel, estimate),
    y = estimate
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Regression coefficient"
  ) +
  theme_minimal()
```

The coefficients indicate how the model relates each marketing variable to revenue while holding the other variables constant.

They should not be interpreted directly as ROI or return on advertising spend.

## Research Question 2: Revenue Prediction

### Question

**How accurately can future revenue be predicted using marketing activity, competitor sales, and seasonality?**

A chronological train/test split is used to evaluate out-of-sample performance.

Observations before 2019 are used for training and observations from 2019 are used for testing.

```{r, echo=FALSE}
train <- df %>%
  filter(
    DATE < as.Date("2019-01-01")
  )

test <- df %>%
  filter(
    DATE >= as.Date("2019-01-01")
  )

train_test_summary <- tibble(
  dataset = c(
    "Training",
    "Test"
  ),
  observations = c(
    nrow(train),
    nrow(test)
  ),
  start_date = c(
    min(train$DATE),
    min(test$DATE)
  ),
  end_date = c(
    max(train$DATE),
    max(test$DATE)
  )
)

train_test_summary
```

The training data contains **163 observations**, while the test data contains **45 observations**.

### Training Model

```{r, echo=FALSE}
model_train <- lm(
  revenue ~
    tv_S +
    ooh_S +
    print_S +
    facebook_S +
    search_S +
    competitor_sales_B +
    month,
  data = train
)

summary(model_train)
```

### Test Set Predictions

```{r, echo=FALSE}
test <- test %>%
  mutate(
    predicted_revenue =
      predict(
        model_train,
        newdata = test
      )
  )
```

### Model Performance

```{r, echo=FALSE}
MAE <- mean(
  abs(
    test$revenue -
      test$predicted_revenue
  )
)

RMSE <- sqrt(
  mean(
    (
      test$revenue -
        test$predicted_revenue
    )^2
  )
)

MAPE <- mean(
  abs(
    (
      test$revenue -
        test$predicted_revenue
    ) /
      test$revenue
  )
) * 100

model_metrics <- tibble(
  metric = c(
    "MAE",
    "RMSE",
    "MAPE"
  ),
  value = c(
    MAE,
    RMSE,
    MAPE
  )
)

model_metrics
```

The model achieves a test-set **MAE of approximately 109,000**, an **RMSE of approximately 212,000**, and a **MAPE of approximately 8.31%**.

The MAPE indicates that the average absolute prediction error is approximately 8.3% of actual revenue during the test period.

### Actual and Predicted Revenue

```{r, echo=FALSE}
ggplot(
  test,
  aes(x = DATE)
) +
  geom_line(aes(y = revenue)) +
  geom_line(
    aes(y = predicted_revenue),
    linetype = "dashed"
  ) +
  scale_y_continuous(labels = comma) +
  scale_x_date(
    breaks = as.Date(c(
      "2019-01-01",
      "2019-04-01",
      "2019-07-01",
      "2019-10-01"
    )),
    labels = c(
      "Jan",
      "Apr",
      "Jul",
      "Oct"
    )
  ) +
  labs(
    x = "Date",
    y = "Revenue"
  ) +
  theme_minimal()
```

The model follows the overall revenue pattern reasonably well, although some individual weeks show larger prediction errors.

## Alternative Model: Log Transformation

A log-transformed regression model was also tested.

The purpose was to examine whether modelling relative changes rather than absolute revenue changes would improve prediction.

```{r, echo=FALSE}
df_log <- df %>%
  mutate(
    log_revenue = log(revenue),
    log_tv = log1p(tv_S),
    log_ooh = log1p(ooh_S),
    log_print = log1p(print_S),
    log_facebook = log1p(facebook_S),
    log_search = log1p(search_S),
    log_competitor = log(competitor_sales_B)
  )

train_log <- df_log %>%
  filter(
    DATE < as.Date("2019-01-01")
  )

test_log <- df_log %>%
  filter(
    DATE >= as.Date("2019-01-01")
  )

model_log <- lm(
  log_revenue ~
    log_tv +
    log_ooh +
    log_print +
    log_facebook +
    log_search +
    log_competitor +
    month,
  data = train_log
)

test_log <- test_log %>%
  mutate(
    predicted_log_revenue =
      predict(
        model_log,
        newdata = test_log
      ),
    predicted_revenue_log =
      exp(predicted_log_revenue)
  )

MAE_log <- mean(
  abs(
    test_log$revenue -
      test_log$predicted_revenue_log
  )
)

RMSE_log <- sqrt(
  mean(
    (
      test_log$revenue -
        test_log$predicted_revenue_log
    )^2
  )
)

MAPE_log <- mean(
  abs(
    (
      test_log$revenue -
        test_log$predicted_revenue_log
    ) /
      test_log$revenue
  )
) * 100
```

The log-transformed model achieved an MAE of approximately **127,000**, RMSE of approximately **209,000**, and MAPE of approximately **8.42%**.

Compared with the baseline model, the log-transformed model has a slightly lower RMSE but worse MAE and MAPE.

The baseline model is therefore preferred for prediction.

## Research Question 3: Delayed Marketing Effects

### Question

**Do marketing activities from previous weeks improve revenue predictions?**

Marketing effects may potentially continue after the week in which spending occurs.

To test this, one- and two-week lags are added for each of the five marketing channels.

```{r, echo=FALSE}
df_lag <- df %>%
  arrange(DATE) %>%
  mutate(
    tv_S_lag1 = lag(tv_S, 1),
    tv_S_lag2 = lag(tv_S, 2),

    ooh_S_lag1 = lag(ooh_S, 1),
    ooh_S_lag2 = lag(ooh_S, 2),

    print_S_lag1 = lag(print_S, 1),
    print_S_lag2 = lag(print_S, 2),

    facebook_S_lag1 = lag(facebook_S, 1),
    facebook_S_lag2 = lag(facebook_S, 2),

    search_S_lag1 = lag(search_S, 1),
    search_S_lag2 = lag(search_S, 2)
  ) %>%
  filter(
    !is.na(tv_S_lag2),
    !is.na(ooh_S_lag2),
    !is.na(print_S_lag2),
    !is.na(facebook_S_lag2),
    !is.na(search_S_lag2)
  )

train_lag <- df_lag %>%
  filter(
    DATE < as.Date("2019-01-01")
  )

test_lag <- df_lag %>%
  filter(
    DATE >= as.Date("2019-01-01")
  )
```

### Lagged Regression Model

```{r, echo=FALSE}
model_lag <- lm(
  revenue ~
    tv_S +
    tv_S_lag1 +
    tv_S_lag2 +

    ooh_S +
    ooh_S_lag1 +
    ooh_S_lag2 +

    print_S +
    print_S_lag1 +
    print_S_lag2 +

    facebook_S +
    facebook_S_lag1 +
    facebook_S_lag2 +

    search_S +
    search_S_lag1 +
    search_S_lag2 +

    competitor_sales_B +
    month,
  data = train_lag
)

test_lag <- test_lag %>%
  mutate(
    predicted_revenue_lag =
      predict(
        model_lag,
        newdata = test_lag
      )
  )

MAE_lag <- mean(
  abs(
    test_lag$revenue -
      test_lag$predicted_revenue_lag
  )
)

RMSE_lag <- sqrt(
  mean(
    (
      test_lag$revenue -
        test_lag$predicted_revenue_lag
    )^2
  )
)

MAPE_lag <- mean(
  abs(
    (
      test_lag$revenue -
        test_lag$predicted_revenue_lag
    ) /
      test_lag$revenue
  )
) * 100
```

### Model Comparison

```{r, echo=FALSE}
model_comparison <- tibble(
  model = c(
    "Baseline linear regression",
    "Log-transformed regression",
    "Lagged marketing regression"
  ),
  MAE = c(
    MAE,
    MAE_log,
    MAE_lag
  ),
  RMSE = c(
    RMSE,
    RMSE_log,
    RMSE_lag
  ),
  MAPE = c(
    MAPE,
    MAPE_log,
    MAPE_lag
  )
)

model_comparison
```

The lagged model achieves an MAE of **136,421**, an RMSE of **225,036**, and a MAPE of **9.85%**.

This is worse than the baseline model, which achieves an MAE of **109,102**, an RMSE of **211,576**, and a MAPE of **8.31%**.

The lagged model therefore does not improve out-of-sample prediction.

### Conclusion

The results suggest that adding one- and two-week marketing lags does not improve revenue prediction in this dataset.

The simpler baseline model is therefore retained.

This is an important modelling result because adding more variables does not automatically improve predictive performance.

## Final Model

Based on the holdout-period results, the baseline linear regression is selected as the final model.

The final model is:

``` text
Revenue ~ TV + OOH + Print + Facebook + Search
          + Competitor Sales + Month
```

The model is preferred because it provides the best overall predictive performance while remaining relatively simple and interpretable.

## Business Insights

### TV

TV has the strongest statistical relationship with revenue among the marketing variables in the final model.

However, the regression coefficient should not be interpreted as a causal return on TV spending.

### Competitor Activity

Competitor sales have a very strong relationship with revenue.

This suggests that broader market conditions are important when evaluating marketing performance.

### Seasonality

Revenue varies substantially across calendar months.

Accounting for seasonality prevents recurring seasonal patterns from being attributed entirely to marketing activity.

### Marketing Spend and Effectiveness

OOH accounts for the largest share of marketing spending, but its coefficient is not statistically significant in the baseline regression.

This illustrates why spending volume alone should not be used to evaluate marketing effectiveness.

### Model Complexity

The lagged model introduced ten additional marketing variables but produced worse predictions.

The results therefore support using the simpler model for this analysis.

## Limitations

### Correlation is Not Causation

The regression estimates associations between variables. It does not establish that changes in marketing spending cause changes in revenue.

### Simulated Data

The dataset is simulated. The results should therefore not be presented as evidence about the real-world effectiveness of any specific marketing channel.

### Limited Model Structure

The model does not explicitly account for:

- adstock
- saturation
- diminishing returns
- carryover effects
- interaction effects

A full Marketing Mix Model could incorporate these effects.

### Multicollinearity

Marketing channels may be correlated with each other and with broader business conditions. This can make individual coefficients difficult to interpret.

### Predictive Rather Than Causal Objective

The main objective of this project is predictive modelling and business interpretation rather than causal inference.

## Conclusion

This project demonstrates a practical regression-based approach to marketing analytics.

The analysis combines exploratory data analysis, seasonality, competitor activity, multiple linear regression, chronological train/test validation, log transformation, and lagged marketing variables.

The final model achieved a test-set MAPE of approximately **8.31%**.

The lagged model performed worse than the simpler baseline model, showing that additional model complexity did not improve predictive performance.

The main takeaway is that marketing analytics should consider marketing channels together with seasonality and broader business drivers rather than relying on simple spend-versus-revenue comparisons.

For a production marketing analytics project, the next step would be to use real business data and develop a more advanced Marketing Mix Modeling approach that accounts for carryover effects, saturation, and causal validation.
